A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2020-02-04 paper/main.tex
fixes the discussion on the top performances and on the evaluation approaches

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.6\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.34.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in APL in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, and Fool Me Once, are
strategies not from the literature but from the APL. DoubleCrosser
is a strategy that makes use of the number of turns because it is set to defect on
the last two rounds. The strategy was expected to not perform as well in
tournaments where the number of turns is not specified, but the strategy did not
perform well in tournaments with noise either. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is strategy specifically designed for noisy environments, however, it ranks
highly in standard tournaments as well.

@@

The top strategies in noisy tournaments are shown in Figure~\ref{fig:noisy_results}. These include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Slow Tit For Two Tats~\cite{axelrodproject}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy
and $e$~\cite{axelrodproject}. Slow Tit For Two Tats is the same strategy as 
Tit For 2 Tats, and at the time of writing this manuscript the
contributors of~\cite{axelrodproject} made a new release where the strategy
has been removed. However, for the purpose of this work the strategy is kept.
The Retaliate and
Limited Retaliate strategies are implemented in APL by the
same contributor. They are strategies designed to defect if the opponent has
tricked them more often than \(x\%\) of the times that they have
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ place are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

From Figure~\ref{fig:noisy_results}, it is evident that the normalised rank
distributions in noisy environments are more variant with higher medians
compared to standard tournaments. The distributions are bimodal. This indicates
that although the top ranked strategies mainly performed well, there are several
tournaments that they ranked in the bottom half. The bimodality of the \(r\)
distributions can be explained by Figure~\ref{fig:noisy_results} which gives the
\(r\) distributions for the top 6 strategies over the noise probability \(p_n\).
Note that for \(p_n=0.5\) a strategy corresponds to a random player and for
\(p_n=1\) a strategy is behaving in the exact opposite way than its design.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_noise.pdf}
    \caption{\(r\) distributions for best performed strategies in noisy tournaments.}
    \label{fig:noisy_results}
\end{figure}

From Figure~\ref{fig:effect_of_noise} it is evident that the strategies highly
ranked in noisy environments did so because of their performance in tournaments
with \(p_n>0.5\). In tournaments with a noise probability lower than \(0.5\) the
strategies performed poorly which is why their \(r\) distributions are bimodal.
If during the data collection a \(p_n\) strictly less 0.5 was considered then the top ranked strategies would be
different. There are a total of 5661 trials where \(p_n<0.5\) and the top ranked
strategies are given in Table~\ref{table:subset_noise_result}. The median ranks
are lower than before and the top spots are mainly overtaken by Meta strategies
which include NMWE deterministic and NMWE Long Memory. The Meta
strategies~\cite{axelrodproject} create a team of strategies for themselves and
choose to play as a member of their team based on their scores against a given
opponent.

@@

% \begin{table}[!htbp]
%     \centering
%     \resizebox{.30\textwidth}{!}{
%     \input{performance_noise_subset_table.tex}}
%     \caption{Top performances in 5661 noisy tournaments where \(p_n<0.5\).}\label{table:subset_noise_result}
% \end{table}

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel Ashlock and Wendy
Ashlock. These strategies have been evolved using reinforcement learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprising that EasyGo and Fool Me Forever which are
strategies that will defect until their opponent defects, then they will cooperate
until the end, ranked $14^{\text{th}}$ and $15^{\text{th}}$. Upon inspection,
it was found that they are actually the same strategy. This was not known to
the authors at the time of data collection. Figure~\ref{fig:probend_results}
verifies that their performance is the same. Both strategies have repeatedly
ranked highly and there are cases for which they were the winners of the
tournament.

The distributions of the normalised rank in probabilistic ending tournaments,
shown in
Figure~\ref{fig:probend_results}, are less variant than those of noisy
tournaments. The medians of the top 15 strategies are lower than 0.1 and the distributions are skewed
towards 0. Though the large difference between the means and the medians
indicates some outliers, the strategies have overall performed well in the
probabilistic ending tournaments that they participated.

\begin{figure}[h!]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_probend.pdf}
    \caption{\(r\) distributions for best performed strategies in probabilistic ending tournaments.}
    \label{fig:probend_results}
\end{figure}

The distributions of \(r\) for the top 6 strategies in probabilistic ending
tournaments over \(p_e\) are given in Figure~\ref{fig:effect_of_probend}.
Figure~\ref{fig:effect_of_probend} shows that the 6 strategies start of with a
high median rank, however, their ranked decreased as the the probability of the
game ending increased and at the point of \(p_e=0.1\) they became the dominant
strategies in their respective tournaments. In essence, what is demonstrated is
that defecting strategies did better when the likelihood of the game ending in
the next turn increased, which is inline with the Folk
Theorem~\cite{Fudenberg2009}. If tournaments where the probability of the game
ending was less than 0.1 were considered then the the top ranked spots are not
dominated by just defecting strategies anymore,
Table~\ref{table:subset_probend_result}. Instead the effective strategies are
now the Meta strategies, trained strategies, Grudger~\cite{axelrodproject} and
Spiteful Tit for Tat~\cite{prison}.

@@

% \begin{table}[!htbp]
%     \centering
%     \resizebox{.30\textwidth}{!}{
%     \input{performance_probend_subset_table.tex}}
%     \caption{Top performances in 1139 probabilistic ending tournaments with \(p_e < 0.1\)}
%     \label{table:subset_probend_result}
% \end{table}

In tournaments with both noise and an unspecified number of turns several of the
top ranked strategies are strategies that were highly ranked in noisy
tournaments. However, strategies from the top ranks in probabilistic ending
tournaments did not rank highly here. Other strategies include $\pi$, $\phi$
which are based on the same approach as $e$. The distributions of \(r\) shown in
Figure~\ref{fig:noisy_probend_results} have the largest median values compared
to the top rank strategies of the other tournament types. A subset of noisy
probabilistic ending tournaments has been considered such that \(p_e < 0.1\) and
\(p_n < 0.5\). The top ranked strategies are given in
Table~\ref{table:subset_probend_noise_result} and it is shown that the Meta
strategies which performed well in noisy tournaments with \(p_n < 0.5\), perform
well once again even the number of turns is not specified. Moreover, several
strategies that did well in probabilistic ending tournaments such as Fortress 3,
Fortress 4, Defector and Better and Better are effective here as well.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.30\textwidth}{!}{
    \input{performance_probend_noise_subset_table.tex}}
    \caption{Top performances in 568 probabilistic ending tournaments with \(p_e < 0.1\) and \(p_n < 0.5\).}
    \label{table:subset_probend_noise_result}
\end{table}

Up till now, the performances of the \numberofstrategies strategies have been evaluated for
individual tournament types. The distributions of \(r\) for the tournament types indicate that
for probabilistic ending and standard tournaments successful strategies do exist.
For these settings, the top 15 strategies have frequently ranked in
the top spots with only a few exceptions. Contrarily, it appears that noise
cause variation in the normalised ranks, and the strategies can always guarantee
a spot in the top ranks.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.35\textwidth}{!}{
    \input{performance_merged_table.tex}}
    \caption{Top performances over all the tournaments}\label{table:overall_results}
\end{table}

@@

@@

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
\resizebox{.8\textwidth}{!}{
    \input{regression_table.tex}}
    \end{center}
\end{table}

To further explore the features that contribute to a strategy's success the
performances are divided into clusters based on whether they were successful or
not. A random forest approach~\cite{breiman2001} is then applied to each
performance to predict the cluster to which it has been assigned to. This allows
for the importance of each feature in the classification to be calculated. In
essence, to calculate which are the important features when one is trying to
predict whether a performance was successful or not.

The performances are clustered into successful and unsuccessful clusters based on
4 different approaches. More specifically:

@@

The random forest method constructs many individual decision trees and the
predictions from all trees are pooled to make the final prediction. The random
forest models are trained on a training set of 70\% of the tournaments results.
The accuracy of each model based on $R^2$ and the number of clusters for each
tournament type (because in the case of Approach 4 it is not
deterministically chosen) are given by Table~\ref{table:accuracy_random_forest}.
The out of the bag error (OOB)~\cite{hastie2005} has also been calculated. The
models fit well, and a high value of both the accuracy measures on the test data
and the OOB error indicate that the model is not over fitting.

@@

The importance that the features of Table~\ref{table:manual_features} had on
each random forest model while the perfomances were clustered based on the
different approaches have been
calculated and are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the features that are highlighted by the importance are inline with
the correlation results. Moreover, the smoothing parameter \(k\) appears to no
have a significant effect either. The most important features based on the
random forest analysis were $C_{r} / C_{median}$ and $C_r / C_{mean}$.

@@

The effect of both these features can be further explored. In
Figure~\ref{fig:mean_median_std} the distributions of \(C_r / C_{\text{mean}}\)
and \(C_r / C_{\text{median}}\) are given for the winners in standard tournaments. A value of \(C_r /
C_{\text{mean}} = 1\) imply that the cooperating ratio of the winner was the
same as the mean/median cooperating ratio of the tournament. In standard tournaments, the mean
for both ratios is 1. Therefore, an effective strategy in standard tournaments
was the mean/median cooperator of its respective tournament. In comparison,
Figure~\ref{fig:mean_median_noisy} shows the distributions of the features for
the winners in noisy tournaments where the mean is at 0.67. Thereupon the winners
cooperated 67\% of the times the mean/median cooperator did. This analysis is
applied to the rest of the tournaments and the distributions are given by
Figures \ref{fig:mean_median_probend}, \ref{fig:mean_median_probend_noisy} and
\ref{fig:mean_median_overall}. In a tournament with noisy and a probabilistic
ending the winners cooperated 60\%, whereas in settings that the type of the
tournament can vary between the types considered in this work the winners
cooperated 67\% of the times the mean or median cooperator did. Finally, in
probabilistic ending tournament it has already been mentioned that defecting
strategies prevail and this result is once again confirmed in this section.

>>>>

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.60\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.34.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in APL in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, and Fool Me Once, are
strategies not from the literature but from the APL. DoubleCrosser
is a strategy that makes use of the number of turns because it is set to defect on
the last two rounds. The strategy was expected to not perform as well in
tournaments where the number of turns is not specified. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is a strategy specifically designed for noisy environments, however, it ranks
highly in standard tournaments as well.

@@

In noisy tournaments the top strategies differ when \(p_n\) is restricted and
when it is not. The \(r\) distributions of the top 6 strategies when \(p_n\in
[0, 1]\) are given by Figure~\ref{fig:effect_of_noise}. It is evident that these
strategies were highly ranked due to the their performance in tournaments with
\(p_n>0.5\), and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly. In comparison, when considering a \(p_n < 0.1\) the
performances of the top strategies are more stable,
Figure~\ref{fig:noise_results}. The top strategies in noisy tournaments with
\(p_n < 0.1\) include DBS, Evolved FSM 16 Noise 05, Evolved ANN 5 Noise 05, PSO
Gambler 2 2 2 Noise 05 and Omega Tit For Tat~\cite{kendall2007iterated} which
are strategies specifically designed for noisy tournaments, strategies such as
BackStabber, DoubleCrosser, Evolved ANN, FSM and HMM which performed well in
standard tournaments as well, and deterministic strategies such as Spiteful Tit For
Tat~\cite{prison}, Level Punisher~\cite{Eckhart2015}, Eugine Nier~\cite{lesswrong}.

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/noise_performance_subset.pdf}
    \caption{$r$ distributions of top 15 strategies in 1151 noisy tournaments with \(p_n < 0.1\).}\label{fig:noise_results}
\end{figure}

In probabilistic ending tournaments it appears that the top ranks are mostly
occupied by defecting strategies and a series of strategies based on finite state
automata introduced by Daniel Ashlock and Wendy Ashlock; Fortress 3, Fortress 4
(both introduced in~\cite{Ashlock2006}), Raider~\cite{Ashlock2014} and
Solution B1~\cite{Ashlock2014}. Values of \(p_e > 0.5\) effectively turn an IPD
match to a one/two shot game, thus it is not surprise that defecting strategies
perform well. From Figure~\ref{fig:effect_of_probend}, which gives the
distributions of \(r\) for the top 6 strategies in probabilistic ending tournaments
over \(p_e\), shows that at the point of \(p_e=0.1\) defecting strategies become
the dominant strategies in their respective tournaments. In essence, what is
demonstrated in Figure~\ref{fig:effect_of_probend} is that defecting strategies
do better when the likelihood of the game ending in the next turn increased,
which is inline with the Folk Theorem~\cite{Fudenberg2009}.

@@

The top spots are not not dominated by just defecting strategies anymore when
\(p_e < 0.1\) is considered, Table~\ref{table:top_performances}. The effective
strategies are now Meta strategies which include Nice Meta Winner, NMWE Finite
Memory, NMWE Deterministic, trained strategies, Grudger~\cite{axelrodproject}
and Spiteful Tit for Tat~\cite{prison}. The Meta
strategies~\cite{axelrodproject} create a team of strategies for themselves and
play as an ensemble or some other combination of the team.
The distributions of the normalised rank in probabilistic ending tournaments,
shown in Figure~\ref{fig:probend_results}, are less variant than those of noisy
tournaments. The medians of the top 15 strategies are lower than 0.1 and the
distributions are skewed towards 0. The strategies have overall performed well
in the probabilistic ending tournaments that they participated.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.6\textwidth]{../images/probend_performance_subset.pdf}
    \caption{\(r\) distributions for best performed strategies in 1139 probabilistic ending
    tournaments with \(p_e < 0.1\).}
    \label{fig:probend_results}
\end{figure}

In tournaments with both noise and an unspecified number of turns several of the
top ranked strategies are strategies that were highly ranked in noisy
tournaments. However, strategies from the top ranks in probabilistic ending
tournaments did not rank highly here.

The distributions of \(r\) shown in Figure~\ref{fig:noisy_probend_results} have
the largest median values compared to the top rank strategies of the other
tournament types. A subset of noisy probabilistic ending tournaments has been
considered such that \(p_e < 0.1\) and \(p_n < 0.1\). The top ranked strategies
are given in Table~\ref{table:subset_probend_noise_result} and it is shown that
the Meta strategies which performed well in noisy tournaments with \(p_n <
0.1\), perform well once again even the number of turns is not specified.
Moreover, several strategies that did well in probabilistic ending tournaments
such as Fortress 4, MEM2 and Spiteful Tit For Tat.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.27\textwidth}{!}{
    \input{performance_probend_noise_subset_table.tex}}
    \caption{Top performances in 117 probabilistic ending tournaments with \(p_e < 0.1\) and \(p_n < 0.1\).}
    \label{table:subset_probend_noise_result}
\end{table}

Up till now, the performances of the \numberofstrategies strategies have been evaluated for
individual tournament types. The distributions of \(r\) for the tournament types indicate that
for probabilistic ending and standard tournaments successful strategies do exist.
For these settings, the top 15 strategies have frequently ranked in
the top spots with only a few exceptions. The behaviour of the winning strategies
in probabilistic ending tournaments vary when \(p_e\) is restricted to 0.1.
For larger values of \(p_e\) the dominant strategies are defecting strategies,
and in comparison for smaller values the dominant strategies are evolved
and strategies based on teams. Similarly, in noisy tournaments the behaviour
of the dominant strategies varied when the probability of noisy was restricted
to 0.1. In noise tournaments with \(p_n \in [0, 1]\) there is much variation
in the normalised ranks, and the strategies can always guarantee a spot in the
top ranks. However, in the restricted case the dominant strategies are strategies
specifically trained for noise tournaments.

@@

\begin{table}[!htbp]
    \centering
    \resizebox{.27\textwidth}{!}{
    \input{performance_merged_table.tex}}
    \caption{Top performances over all the tournaments}\label{table:overall_results}
\end{table}

@@

\textbf{Evaluation based on correlation.}

@@

\textbf{Evaluation based on clustering and random forest.}

A second approach for evaluating the features importance in a strategy's success
is a combination of clustering and a random forest algorithm. Initially,
the strategies performances are clustered four different time based on the
following approaches:

@@

Once the performances have been assigned to a cluster for each approach a
random forest algorithm~\cite{breiman2001} is applied. The problem is a supervised
problem where the random forest algorithm predicts the cluster to which a performance
has been assigned to using the features of Table~\ref{table:manual_features}.
The random
forest models are trained on a training set of 70\% of the tournaments results.
The accuracy of each model based on $R^2$ and the number of clusters for each
tournament type (because in the case of Approach 4 it is not
deterministically chosen) are given by Table~\ref{table:accuracy_random_forest}.
The out of the bag error (OOB)~\cite{hastie2005} has also been calculated. The
models fit well, and a high value of both the accuracy measures on the test data
and the OOB error indicate that the model is not over fitting.

@@

The importance that the features of Table~\ref{table:manual_features} had on
each random forest model while the performances were clustered based on the
different approaches have been
calculated and are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the features that are highlighted by the importance are inline with
the correlation results. Moreover, the smoothing parameter \(k\) appears to no
have a significant effect either. The most important features based on the
random forest analysis were $C_{r} / C_{median}$ and $C_r / C_{mean}$.

@@

\textbf{Evaluation based on linear regression.}

A multivariate linear regression is also fitted to model the relationship of the
features and the normalised rank. The \(p-\)values of each feature as well as
their coefficients are given by Table~\ref{table:linear_regression}. The
coefficient of determination \(R\) squared fo each model is also report in
Table~\ref{table:linear_regression}. Based on \(R\) square the model explain 90\% of
the variability of the response data around its mean.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h]
    \begin{center}
\resizebox{.8\textwidth}{!}{
    \input{regression_table.tex}}
    \end{center}
    \caption{Multivariate linear regression with \(r\) as the depended variable results.}
    \label{table:linear_regression}
\end{table}

The results of the linear regression verify the results of the previous approaches.

The features $C_{r} / C_{median}$ and $C_r / C_{mean}$ have been repeatably
highlighted as importance feature in explaining the variability of the performances.
Thus the effect of these tow measures is further explored.In
Figure~\ref{fig:mean_median_std} the distributions of \(C_r / C_{\text{mean}}\)
and \(C_r / C_{\text{median}}\) are given for the winners in standard tournaments. A value of \(C_r /
C_{\text{mean}} = 1\) imply that the cooperating ratio of the winner was the
same as the mean/median cooperating ratio of the tournament. In standard tournaments, the mean
for both ratios is 1. Therefore, an effective strategy in standard tournaments
was the mean/median cooperator of its respective tournament. In comparison,
Figure~\ref{fig:mean_median_noisy} shows the distributions of the features for
the winners in noisy tournaments where the mean is at 0.67. Thereupon the winners
cooperated 67\% of the times the mean/median cooperator did. This analysis is
applied to the rest of the tournaments and the distributions are given by
Figures \ref{fig:mean_median_probend}, \ref{fig:mean_median_probend_noisy} and
\ref{fig:mean_median_overall}. In a tournament with noisy and a probabilistic
ending the winners cooperated 60\%, whereas in settings that the type of the
tournament can vary between the types considered in this work the winners
cooperated 67\% of the times the mean or median cooperator did. Finally, in
probabilistic ending tournament it has already been mentioned that defecting
strategies prevail and this result is once again confirmed in this section.
